Climate predictor selection and variable reduction methods influence MaxEnt model performance and predictions of climatic suitability

### 2023-06-05 manuscript/manuscript.qmd
Bulk dump of text to intro and methods

---
title: Climate predictor selection and variable reduction methods influence MaxEnt model performance and predictions of climatic suitability 
author:
  - name: Clarke van Steenderen
    email: vsteenderen@gmail.com
    affiliations: 
        - id: some-tech
          name: Center for Biological Control, Department of Zoology and Entomology
          department: Rhodes University
          address: 
          city: Makhanda
          state: South Africa
          postal-code: 6140
    attributes:
        corresponding: false
  - name: Guy F. Sutton
    email: g.sutton@ru.ac.za
    affiliations:
        - id: another-u
          name: Center for Biological Control, Department of Zoology and Entomology
          department: Rhodes University
          address: 
          city: Makhanda
          state: South Africa
          postal-code: 6140
    attributes:
        corresponding: true
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - Species distribution model
  - Variable selection
  - Ecological model
  - WORLDCLIM
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Modelling
      formatting: preprint
      model: 3p
      layout: twocolumn
      cite-style: authoryear
      highlights:
        - Highlight 1
        - Highlight 2 
        - Highlight 3
---

@@

Ecological models are important tools to aid the development and implementation of environmental policies and management programmes [@Addison2013; @Schuwirth2019; @Sutton2022]. These models are used for conservation planning [@Guisan2013], predicting the establishment and spread of invasive species [@Martin2020], implementing biological control programmes [@Sutton2019b; @Mukherjee2021], and forecasting species responses to environmental change [@Bocedi2014], amongst other applications. Species distribution models (SDM’s) are an example of ecological models that have become increasingly popular in recent years [@Elith2009]. SDM’s typically take the form of correlative or mechanistic models that correlate species presence/absences (or pseudo-absences) to environmental covariates to identify suitable climatic conditions for the study taxon [@Elith2011]. The Maximum Entropy species distribution model (hereafter ‘MaxEnt’) is amongst the most popular methods for climate modelling studies and has been shown to perform well compared to alternative modelling algorithms [@Wisz2008; @Phillips2017]. It uses maximum entropy to distinguish between environmental conditions where the focal taxon is present from environmental conditions at sites without confirmed presence records for the taxon [@Elith2011].

@@

**Next paragraphs**

- Review current methods used to select variables, e.g. R2, VIF, PCA, expert opinion, and most recently, automated selection / model-based selection.

- Aims of this paper are to show how variable selection can influence model outputs and inferences.

@@

### 2.1. Species occurrence records

@@

Spatial autocorrelation is an important factor that may affect model outputs. Filtering of species occurrence data may limit the inherent biases in the data and improve model quality (Veloz, 2009). To avoid pseudo-replication, only one occurrencenrecord per 2.5 min grid cell was used for model calibration. Species occurrence datasets were thinned using the ‘*spThin*’ package [@Aiello-Lammens2015], and spatial autocorrelation analyses were performed using the ‘*ecospat*’ package [@DiCola2017]. Eighty-eight occurrence records were retained to calibrate native-range models. Spatial thinning was also performed for the invaded range occurrences, however, there was no evidence for spatial autocorrelation. As such, no occurrence records were thinned for the invaded-range dataset.

### 2.2. Environmental predictors

@@

### 2.3. Model calibration

@@

Given that MaxEnt is a presence/pseudo-absence modelling algorithm, model calibration requires a user-defined geographic background to sample the climate of representative grid cells where the focal species is assumed to be absent (i.e., background points or pseudo-absences). Background definition can have a significant effect on model output (VanDerWal et al., 2009). The background should ideally represent the geographic areas available to the focal species, omitting areas where species absence is due to historical factors, dispersal constraints and/or biotic interactions (Sanin and Anderson, 2018). Following Webber et al. (2011), we defined the model background using the Koppen-Geiger climate classification system (available at: http://koeppen-geiger.vu-w ien.ac.at). Only Koppen-Geiger climate zones that contained at least one native-range occurrence record for D. rubiformis in Australia were used as the background area from which background points were drawn for model calibration (Fig. 1a). The Koppen-Geiger climate zones were intersected with the occurrence records using the ‘ raster’ R package (Hijmans, 2022). We randomly sampled 10 000 points (the default number used for Maxent; Merow et al. (2013) ) from within this background definition using the ‘dismo’ R package (Hijmans et al., 2021).

MaxEnt models were parameterised with default settings for multiple parameters, including: convergence = 105, maximum number of iterations = 500 and prevalence = 0.5. The ‘fade by clamping’ option was selected to prevent extrapolation well outside the range of climatic values in the model training area (Philips et al., 2017 ). Model predictions were obtained using the ‘logistic output’ to create continuous climatic suitability raster layers scaled between 0 (climatically unsuitable) and 1 (climatically suitable).

### 2.4. Model evaluation

@@

### 2.5. Model visualisation

@@

All modelling and statistical analyses were conducted in R ver. 4.0.3 (R Core Team, 2020). All values presented in text are presentedas mean ± standard error, unless otherwise stated. A standardised ODMAP methods protocol (Overview, Data, Model, Assessment, and Prediction) has been completed for this study and can be found in Supplementary File S1. ODMAP standardises the reporting of SDM modelling studies to improve transparency and reproducibility (Zurell et al., 2020).

>>>>

---
title: Climate predictor selection and variable reduction methods influence MaxEnt model performance and predictions of climatic suitability 
author:
  - name: Clarke van Steenderen
    email: vsteenderen@gmail.com
    affiliations: 
        - id: some-tech
          name: Center for Biological Control, Department of Zoology and Entomology
          department: Rhodes University
          address: 
          city: Makhanda
          state: South Africa
          postal-code: 6140
    attributes:
        corresponding: false
  - name: Guy F. Sutton
    email: g.sutton@ru.ac.za
    affiliations:
        - id: another-u
          name: Center for Biological Control, Department of Zoology and Entomology
          department: Rhodes University
          address: 
          city: Makhanda
          state: South Africa
          postal-code: 6140
    attributes:
        corresponding: true
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - Species distribution model
  - Variable selection
  - Ecological model
  - WORLDCLIM
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Modelling
      formatting: preprint
      model: 3p
      layout: onecolumn
      cite-style: authoryear
      highlights:
        - Highlight 1
        - Highlight 2 
        - Highlight 3
---

@@

Ecological models are important tools to aid the development and implementation of environmental policies and management programmes [@Addison2013; @Schuwirth2019; @Sutton2022]. These models are used for conservation planning [@Guisan2013], predicting the establishment and spread of invasive species [@Martin2020], implementing biological control programmes [@Sutton2019b; @Mukherjee2021], and forecasting species responses to environmental change [@Bocedi2014], amongst other applications. Species distribution models (SDM’s) are an example of ecological models that have become increasingly popular in recent years [@Elith2009]. SDM’s typically take the form of correlative or mechanistic models that correlate species presence/absences (or pseudo-absences) to environmental covariates to identify suitable climatic conditions for the study taxon [@Elith2011]. The Maximum Entropy species distribution model (hereafter ‘*MaxEnt*’) is amongst the most popular methods for climate modelling studies and has been shown to perform well compared to alternative modelling algorithms [@Wisz2008; @Phillips2017]. It uses maximum entropy to distinguish between environmental conditions where the focal taxon is present from environmental conditions at sites without confirmed presence records for the taxon [@Elith2011].

@@

The methods used for selecting covariates for SDM studies typically fall under four categories: (1) creating a set of uncorrelated covariates from a larger expert-chosen group of 15-30 covariates, using correlation analyses and/or variance inflation factors (VIF's) [@Dormann2013], (2) using principal component analysis (PCA) to reduce dimensionality of a set of covariates [@Junior2018], (3) using expert-opinion to select covariates that may be relevant based on the knowledge of the focal species biology, physiology and ecological requirements [@Petitpierre2017; @Scherrer2019], and more recently (4) automated statistical model-based covariate selection techniques, such as '*embedded methods*', e.g. RIDGE and LASSO [@Guisan2002; @Guyon2003; @Saeys2007; @Adde2023]. For a more detailed overview of the different variable selection methods, we direct readers to several recent reviews on this topic [@Fois2018; @Fourcade2018; @Melo-Merino2020].

While there are guidelines for how to implement these methods individually, e.g. using an |*r*| > 0.70 or a VIF > 2-5 as a threshold for removing significantly correlated predictors [@Dormann2013], there is currently no consensus on which method to use for a particular study (but see @Adde2023). As such, we aimed to demonstrate the variation in climatic suitability raster projections and predictive accuracy of MaxEnt models calibrated using the different covariate selection methods, and discuss the potential implications for model inference and possible management programmes based on these inferences. To demonstrate, we modelled the potential climatic suitability (i.e. the climatic similarity relative to the climate space occupied in the model calibration region) for the invasive insect pest *Diaphorina citri* Kuwayama (Asian citrus psyllid) (Hemiptera: Psyllidae), that is native to Asia but has become invasive in many non-native regions around the world, such as the USA, Brazil and Africa ( **Insert citations**, **Insert a map of global distribution?** ).

- Clarke: Can you please insert 1-2 sentences about the impact of ACP and its association with HLB?

@@

### Species occurrence records

@@

Spatial autocorrelation is an important factor that may affect model outputs. Filtering of species occurrence data may limit the inherent biases in the data and improve model quality [@Veloz2009]. To avoid pseudo-replication, only one occurrence record per 2.5 minute grid cell was used for model calibration. Species occurrence datasets were thinned using the ‘*spThin*’ package [@Aiello-Lammens2015], and spatial autocorrelation analyses were performed using the ‘*ecospat*’ package [@DiCola2017]. Eighty-eight occurrence records were retained to calibrate native-range models. Spatial thinning was also performed for the invaded range occurrences, however, there was no evidence for spatial autocorrelation. As such, no occurrence records were thinned for the invaded-range dataset.

### Environmental predictors

@@

### Model calibration

@@

Given that MaxEnt is a presence/pseudo-absence modelling algorithm, model calibration requires a user-defined geographic background to sample the climate of representative grid cells where the focal species is assumed to be absent (i.e., background points or pseudo-absences). Background definition can have a significant effect on model output [@VanDerWal2009]. The background should ideally represent the geographic areas available to the focal species, omitting areas where species absence is due to historical factors, dispersal constraints and/or biotic interactions [@Sanin2018]. We defined the model background using the Koppen-Geiger climate classification system (available at: http://koeppen-geiger.vu-wien.ac.at) [@Webber2011]. Only Koppen-Geiger climate zones that contained at least one native-range occurrence record for D. rubiformis in Australia were used as the background area from which background points were drawn for model calibration (Fig. 1a).

* Need to update the below with appropriate references to 'terra' package

The Koppen-Geiger climate zones were intersected with the occurrence records using the ‘ raster’ R package (Hijmans, 2022). We randomly sampled 10 000 points (the default number used for Maxent; Merow et al. (2013) ) from within this background definition using the ‘dismo’ R package (Hijmans et al., 2021).

MaxEnt models were specified with default settings for multiple parameters, including: convergence = 105, maximum number of iterations = 500 and prevalence = 0.5. The ‘fade by clamping’ option was selected to prevent extrapolation well outside the range of climatic values in the model training area [@Phillips2017]. Model predictions were obtained using the ‘logistic output’ to create continuous climatic suitability raster layers scaled between 0 (climatically unsuitable) and 1 (climatically suitable).

### Model evaluation

@@

### Model visualisation

@@

All modelling and statistical analyses were conducted in *R* ver. 4.3.0 [@Team2023]. All values presented in text are presented as mean ± standard error, unless otherwise stated. A standardised ODMAP methods protocol (Overview, Data, Model, Assessment, and Prediction) has been completed for this study and can be found in Supplementary File S1. ODMAP standardises the reporting of SDM modelling studies to improve transparency and reproducibility (Zurell et al., 2020).

### 2023-06-05 manuscript/manuscript.tex
Bulk dump of text to intro and methods

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p,
  twocolumn]{elsarticle}

@@

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\usepackage{float}
\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother
\journal{Ecological Modelling}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Climate predictor selection and variable reduction methods influence MaxEnt model performance and predictions of climatic suitability},
  pdfauthor={Clarke van Steenderen; Guy F. Sutton},
  pdfkeywords={keyword1, keyword2},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    keyword1 \sep 
    keyword2
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, interior hidden, sharp corners, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Ecological models are important tools to aid the development and
implementation of environmental policies and management programmes
\citep{Addison2013, Schuwirth2019, Sutton2022}. These models are used
for conservation planning \citep{Guisan2013}, predicting the
establishment and spread of invasive species \citep{Martin2020},
implementing biological control programmes
\citep{Sutton2019b, Mukherjee2021}, and forecasting species responses to
environmental change \citep{Bocedi2014}, amongst other applications.
Species distribution models (SDM's) are an example of ecological models
that have become increasingly popular in recent years \citep{Elith2009}.
SDM's typically take the form of correlative or mechanistic models that
correlate species presence/absences (or pseudo-absences) to
environmental covariates to identify suitable climatic conditions for
the study taxon \citep{Elith2011}. The Maximum Entropy species
distribution model (hereafter `MaxEnt') is amongst the most popular
methods for climate modelling studies and has been shown to perform well
compared to alternative modelling algorithms
\citep{Wisz2008, Phillips2017}. It uses maximum entropy to distinguish
between environmental conditions where the focal taxon is present from
environmental conditions at sites without confirmed presence records for
the taxon \citep{Elith2011}.

@@

\textbf{Next paragraphs}

\begin{itemize}
\item
  Review current methods used to select variables, e.g.~R2, VIF, PCA,
  expert opinion, and most recently, automated selection / model-based
  selection.
\item
  Aims of this paper are to show how variable selection can influence
  model outputs and inferences.
\end{itemize}

@@

\hypertarget{species-occurrence-records}{%
\subsubsection{2.1. Species occurrence
records}\label{species-occurrence-records}}

\hypertarget{environmental-predictors}{%
\subsubsection{2.2. Environmental
predictors}\label{environmental-predictors}}

Climate data were obtained by downloading the standard set of 19
bioclimatic variables from the WorldClim ver. 2.1 database (Fick and
Hijmans, 2017) (data available at: www.worldclim.org/download. html).
This dataset is representative of annual and seasonal means and
variation of temperature and precipitation metrics averaged over the
1950--2000 time period (current climate) at a 2.5 min resolution. These
variables have been shown to effectively model the climatic suitability
for non-native insects (e.g., Trethowan et al., 2011).

\hypertarget{model-calibration}{%
\subsubsection{2.3. Model calibration}\label{model-calibration}}

MaxEnt (ver. 3.4.3) was implemented in the ` dismo' R package (Hijmans
et al., 2021 ).

Given that MaxEnt is a presence/pseudo-absence modelling algorithm,
model calibration requires a user-defined geographic background to
sample the climate of representative grid cells where the focal species
is assumed to be absent (i.e., background points or pseudo-absences).
Background definition can have a significant effect on model output
(VanDerWal et al., 2009). The background should ideally represent the
geographic areas available to the focal species, omitting areas where
species absence is due to historical factors, dispersal constraints
and/or biotic interactions (Sanin and Anderson, 2018). Following Webber
et al.~(2011), we defined the model background using the Koppen-Geiger
climate classification system (available at: http://koeppen-geiger.vu-w
ien.ac.at). Only Koppen-Geiger climate zones that contained at least one
native-range occurrence record for D. rubiformis in Australia were used
as the background area from which background points were drawn for model
calibration (Fig. 1a). The Koppen-Geiger climate zones were intersected
with the occurrence records using the ` raster' R package (Hijmans,
2022). We randomly sampled 10 000 points (the default number used for
Maxent; Merow et al.~(2013) ) from within this background definition
using the `dismo' R package (Hijmans et al., 2021).

MaxEnt models were parameterised with default settings for multiple
parameters, including: convergence = 105, maximum number of iterations =
500 and prevalence = 0.5. The `fade by clamping' option was selected to
prevent extrapolation well outside the range of climatic values in the
model training area (Philips et al., 2017 ). Model predictions were
obtained using the `logistic output' to create continuous climatic
suitability raster layers scaled between 0 (climatically unsuitable) and
1 (climatically suitable).

\hypertarget{model-evaluation}{%
\subsubsection{2.4. Model evaluation}\label{model-evaluation}}

@@

\hypertarget{model-visualisation}{%
\subsubsection{2.5. Model visualisation}\label{model-visualisation}}

@@

All modelling and statistical analyses were conducted in R ver. 4.0.3 (R
Core Team, 2020). All values presented in text are presentedas mean ±
standard error, unless otherwise stated. A standardised ODMAP methods
protocol (Overview, Data, Model, Assessment, and Prediction) has been
completed for this study and can be found in Supplementary File S1.
ODMAP standardises the reporting of SDM modelling studies to improve
transparency and reproducibility (Zurell et al., 2020).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p,
  onecolumn]{elsarticle}

@@

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Ecological Modelling}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Climate predictor selection and variable reduction methods influence MaxEnt model performance and predictions of climatic suitability},
  pdfauthor={Clarke van Steenderen; Guy F. Sutton},
  pdfkeywords={Species distribution model, Variable
selection, Ecological model, WORLDCLIM},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    Species distribution model \sep Variable selection \sep Ecological
model \sep 
    WORLDCLIM
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, boxrule=0pt, enhanced, interior hidden]}{\end{tcolorbox}}\fi

@@

Ecological models are important tools to aid the development and
implementation of environmental policies and management programmes
\citep{Addison2013, Schuwirth2019, Sutton2022}. These models are used
for conservation planning \citep{Guisan2013}, predicting the
establishment and spread of invasive species \citep{Martin2020},
implementing biological control programmes
\citep{Sutton2019b, Mukherjee2021}, and forecasting species responses to
environmental change \citep{Bocedi2014}, amongst other applications.
Species distribution models (SDM's) are an example of ecological models
that have become increasingly popular in recent years \citep{Elith2009}.
SDM's typically take the form of correlative or mechanistic models that
correlate species presence/absences (or pseudo-absences) to
environmental covariates to identify suitable climatic conditions for
the study taxon \citep{Elith2011}. The Maximum Entropy species
distribution model (hereafter `\emph{MaxEnt}') is amongst the most
popular methods for climate modelling studies and has been shown to
perform well compared to alternative modelling algorithms
\citep{Wisz2008, Phillips2017}. It uses maximum entropy to distinguish
between environmental conditions where the focal taxon is present from
environmental conditions at sites without confirmed presence records for
the taxon \citep{Elith2011}.

@@

The methods used for selecting covariates for SDM studies typically fall
under four categories: (1) creating a set of uncorrelated covariates
from a larger expert-chosen group of 15-30 covariates, using correlation
analyses and/or variance inflation factors (VIF's) \citep{Dormann2013},
(2) using principal component analysis (PCA) to reduce dimensionality of
a set of covariates \citep{Junior2018}, (3) using expert-opinion to
select covariates that may be relevant based on the knowledge of the
focal species biology, physiology and ecological requirements
\citep{Petitpierre2017, Scherrer2019}, and more recently (4) automated
statistical model-based covariate selection techniques, such as
`\emph{embedded methods}', e.g.~RIDGE and LASSO
\citep{Guisan2002, Guyon2003, Saeys2007, Adde2023}. For a more detailed
overview of the different variable selection methods, we direct readers
to several recent reviews on this topic
\citep{Fois2018, Fourcade2018, Melo-Merino2020}.

While there are guidelines for how to implement these methods
individually, e.g.~using an \textbar{}\emph{r}\textbar{} \textgreater{}
0.70 or a VIF \textgreater{} 2-5 as a threshold for removing
significantly correlated predictors \citep{Dormann2013}, there is
currently no consensus on which method to use for a particular study
(but see \citet{Adde2023}). As such, we aimed to demonstrate the
variation in climatic suitability raster projections and predictive
accuracy of MaxEnt models calibrated using the different covariate
selection methods, and discuss the potential implications for model
inference and possible management programmes based on these inferences.
To demonstrate, we modelled the potential climatic suitability (i.e.~the
climatic similarity relative to the climate space occupied in the model
calibration region) for the invasive insect pest \emph{Diaphorina citri}
Kuwayama (Asian citrus psyllid) (Hemiptera: Psyllidae), that is native
to Asia but has become invasive in many non-native regions around the
world, such as the USA, Brazil and Africa ( \textbf{Insert citations},
\textbf{Insert a map of global distribution?} ).

\begin{itemize}
\tightlist
\item
  Clarke: Can you please insert 1-2 sentences about the impact of ACP
  and its association with HLB?
\end{itemize}

@@

\hypertarget{species-occurrence-records}{%
\subsubsection{Species occurrence
records}\label{species-occurrence-records}}

\begin{itemize}
\tightlist
\item
  Clarke to add information on where we sourced GPS records - link to
  key papers, maybe a supplementary table (if required).
\item
  How many records do we have?
\end{itemize}

Spatial autocorrelation is an important factor that may affect model
outputs. Filtering of species occurrence data may limit the inherent
biases in the data and improve model quality \citep{Veloz2009}. To avoid
pseudo-replication, only one occurrence record per 2.5 minute grid cell
was used for model calibration. Species occurrence datasets were thinned
using the `\emph{spThin}' package \citep{Aiello-Lammens2015}, and
spatial autocorrelation analyses were performed using the
`\emph{ecospat}' package \citep{DiCola2017}. Eighty-eight occurrence
records were retained to calibrate native-range models. Spatial thinning
was also performed for the invaded range occurrences, however, there was
no evidence for spatial autocorrelation. As such, no occurrence records
were thinned for the invaded-range dataset.

\hypertarget{environmental-predictors}{%
\subsubsection{Environmental
predictors}\label{environmental-predictors}}

Climate data were obtained by downloading the standard set of 19
bioclimatic variables from the WorldClim ver. 2.1 database
\citep{Fick2017} (data available at: www.worldclim.org/download.html).
This dataset is representative of annual and seasonal means and
variation of temperature and precipitation metrics averaged over the
1950--2000 time period (current climate) at a 2.5 minute resolution.
These variables have been shown to effectively model the climatic
suitability for non-native insects (e.g., \citet{Sutton2022}).

\hypertarget{model-calibration}{%
\subsubsection{Model calibration}\label{model-calibration}}

MaxEnt (ver. 3.4.3) was implemented in the `\emph{dismo}' R package
(\citet{Hijmans2017}).

Given that MaxEnt is a presence/pseudo-absence modelling algorithm,
model calibration requires a user-defined geographic background to
sample the climate of representative grid cells where the focal species
is assumed to be absent (i.e., background points or pseudo-absences).
Background definition can have a significant effect on model output
\citep{VanDerWal2009}. The background should ideally represent the
geographic areas available to the focal species, omitting areas where
species absence is due to historical factors, dispersal constraints
and/or biotic interactions \citep{Sanin2018}. We defined the model
background using the Koppen-Geiger climate classification system
(available at: http://koeppen-geiger.vu-wien.ac.at) \citep{Webber2011}.
Only Koppen-Geiger climate zones that contained at least one
native-range occurrence record for D. rubiformis in Australia were used
as the background area from which background points were drawn for model
calibration (Fig. 1a).

\begin{itemize}
\tightlist
\item
  Need to update the below with appropriate references to `terra'
  package
\end{itemize}

The Koppen-Geiger climate zones were intersected with the occurrence
records using the ` raster' R package (Hijmans, 2022). We randomly
sampled 10 000 points (the default number used for Maxent; Merow et
al.~(2013) ) from within this background definition using the `dismo' R
package (Hijmans et al., 2021).

MaxEnt models were specified with default settings for multiple
parameters, including: convergence = 105, maximum number of iterations =
500 and prevalence = 0.5. The `fade by clamping' option was selected to
prevent extrapolation well outside the range of climatic values in the
model training area \citep{Phillips2017}. Model predictions were
obtained using the `logistic output' to create continuous climatic
suitability raster layers scaled between 0 (climatically unsuitable) and
1 (climatically suitable).

\hypertarget{model-evaluation}{%
\subsubsection{Model evaluation}\label{model-evaluation}}

@@

\hypertarget{model-visualisation}{%
\subsubsection{Model visualisation}\label{model-visualisation}}

@@

All modelling and statistical analyses were conducted in \emph{R} ver.
4.3.0 \citep{Team2023}. All values presented in text are presented as
mean ± standard error, unless otherwise stated. A standardised ODMAP
methods protocol (Overview, Data, Model, Assessment, and Prediction) has
been completed for this study and can be found in Supplementary File S1.
ODMAP standardises the reporting of SDM modelling studies to improve
transparency and reproducibility (Zurell et al., 2020).
